---
Title: Red Wine - Data Exploration
Author: Karunakaran TK
output: rmarkdown::github_document
---
Red Wine - Data Exploration
========================================================

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(GGally)
library(scales)
library(lattice)
library(MASS)
library(gridExtra)
library(dplyr)
library(reshape)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
redwine <- read.csv('wineQualityReds.csv')
```

## Introduction to the dataset
This dataset has 1599 observations realted to the red variant of the Portuguese 
"Vinho Verde" wine. Variables are the physiochemical and sensory attributes
of this variant of red wine. There are no missing values. Information on the 
13 variables in this dataset can be seen below:

```{r}
str(redwine)
```


# Univariate Plots & Analysis

The primary output variable of this dataset is the 'quality' variable. This
looks like a good place to start the exploration of this dataset. Let us look
at a histogram of the quaity variable

```{r echo=FALSE, Univariate_Plots-Plot1}
ggplot(aes(x = quality), data = redwine) +
  geom_bar(fill="#AEC7E8", stat='count') +
  labs(x='Wine Quality', y='Count') +
  ggtitle("Distribution of Wine Quality") +
  theme(panel.background = element_blank())
```


```{r echo=FALSE, Univariate_Plots-Plot2}
#plotting the fixed.acidity distribution
ggplot(aes(x = fixed.acidity), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.5) +
  ggtitle("Fixed Acidity") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

Fixed acidity is the non volatile acids present in wine. Wines in this dataset
has an average fixed acidity of 8.32 g/dm^3. We can see from the above
histogram that the distribution is a bit negatively skewed, indicating the
presence of a few outliers with high amounts of fixed acidity.

```{r echo=FALSE, Univariate_Plots-Plot3}
#plotting the volatile.acidity distribution
 ggplot(aes(x = volatile.acidity), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.15) +
  ggtitle("Volatile Acidity") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

The volatile acidity of the wine is the amount of acetic acid in wine, which at 
too high of levels can lead to an unpleasant, vinegar taste. Wines in this 
dataset have an average volaitle acidity of 0.52 g/dm^3. From the distribution, 
we see that like fixed acidity, the volatile acidity is also negatively skewed, 
with a few wines having high volatile acidity (outliers). We suspect that these 
wines might be of low quality.

```{r echo=FALSE, Univariate_Plots-Plot4}
ggplot(aes(x = citric.acid), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.05) +
  ggtitle("Citric Acidity") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

Citric acid is found in small quantities. It can add freshness and flavor
to the wine. We see that there are less number of wines with higher levels of
citric acid. On average, wines in this dataset have 0.27 g/dm^3 of citric
acid.

```{r echo=FALSE, Univariate_Plots-Plot5}
ggplot(aes(x = residual.sugar), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.5) +
  ggtitle("Residual Sugar") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

Residual sugar, which is the amount of sugar that remains after fermentaiton
stops, has a heavily skewed long tailed distribution with many outliers. This 
is more clearly seen in the below boxplot

```{r echo=FALSE, Univariate_Plots-Plot6}
boxplot(redwine$residual.sugar, horizontal = T, frame.plot = FALSE,
        main='Residual Sugar', pch=21, cex = .4)
axis(side=1, col='white')
```

```{r echo=FALSE, Univariate_Plots-Plot7}
ggplot(aes(x = chlorides), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.03) +
  ggtitle("Chlorides") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

Chlorides, which is the amount of salt in the wine, is also a heavily skewed
distribution, similar to residual sugar. There are many outliers. On average, 
wines in this dataset have 0.08 g/dm^3 of cholrides in them. As we can see from
the plot, there are outliers that go as high as 0.6 g/dm^3 of chlorides.

```{r echo=FALSE, Univariate_Plots-Plot8}
ggplot(aes(x = free.sulfur.dioxide), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 4) +
  ggtitle("Free Sulpher Dioxide") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

Free sulfur dioxide: the free form of SO2 exists in equilibrium between
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine. There are more wines in the dataset with low
levels of free sulfur dioxide, than those with more. On average, wines contain
15.87 mg/dm^3 of free sulfur dioxide

```{r echo=FALSE, Univariate_Plots-Plot9}
ggplot(aes(x = total.sulfur.dioxide), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 10) +
  ggtitle("Total Sulpher Dioxide") +
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank(),
  axis.title.x=element_blank())
```

This is the amount of free and bound forms of sulfur dioxide. Similar to free
sulfur dioxide, the distribution of total sulfur dioxide is also positively
skewed with few wines with extreme values of total sulfur dioxide. There are
two large outliers in this dataset as can be seen from the below box plot.

```{r echo=FALSE, Univariate_Plots-Plot10}
boxplot(redwine$total.sulfur.dioxide, horizontal = T, frame.plot = FALSE,
        main='Total Sulfur Dioxide', pch=21, cex = .4)
axis(side=1, col='white')
```

```{r echo=FALSE, Univariate_Plots-Plot11}
ggplot(aes(x = density), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.001) +
  ggtitle("Density") +
  xlab('Density (g/cm^3)')+
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank())
```

Density of water in the wine is one of the few normally distributed variables
in this dataset. The median and mean is roughly the same (0.99 g/cm^3)

```{r echo=FALSE, Univariate_Plots-Plot12}
ggplot(aes(x = pH), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.05) +
  ggtitle("pH") +
  xlab('pH')+
  theme(axis.text.y= element_text(size=7), 
  panel.background = element_blank())
```

Another variable in the dataset which is normally distributed. pH describes 
how acidic or basic the wine is on a scale of 0(very acidic) to 14 (very basic).
Most wine fall in the 3-4 range.

```{r echo=FALSE, Univariate_Plots-Plot13}
ggplot(aes(x = sulphates), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.1) +
  ggtitle("Sulphates") +
  xlab('Sulphates (g/dm^3)')+
  theme(axis.text.y= element_text(size=7), 
    panel.background = element_blank())
```

Sulphates refer to additives that can contribute to sulfur dioxide in the wine.
The distribution of sulphates is positively skewed with a few outliers. The 
average amout of sulphates is 0.6 g/dm^3.

```{r echo=FALSE, Univariate_Plots-Plot14}
ggplot(aes(x = alcohol), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 1) +
  ggtitle("Alcohol") +
  xlab('Alcohol (%)')+
  theme(axis.text.y= element_text(size=7), 
    panel.background = element_blank())
```

There are less number of wine with high % of alcohol content in them. Average
alcohol content is around 10.5%. 


We can construct a new variable called total acidity, which is the sum of fixed
and volatile acidity. 

```{r echo=FALSE, Univariate_Plots-Plot15}
redwine$total.acidity <- redwine$fixed.acidity+redwine$volatile.acidity

ggplot(aes(x = total.acidity), data = redwine)+
  geom_histogram(colour="#8295AE", fill="#1f77b4",binwidth = 0.5) +
  ggtitle("Total Acidity") +
  xlab('Total Acidity (g/dm^3)')+
  theme(axis.text.y= element_text(size=7), 
    panel.background = element_blank())
```

The distribution of total acidity looks very similar to that of fixed acidity,
since the volatile acidity in all wines are very small.

### Structure of your dataset?
The dataset has 1599 observations and 13 variables. All variables are numeric.
Most of the variables are positively skewed, excelpt pH, density and volatile
acidity, which appear to be normally distributed. 

### Main feature(s) of interest in your dataset?
There are 11 input variables and one output variables which is the quality of
the wine. The 11 input variables which relate to acidity, residual sugar, 
chlorides, free and total sulpher dioxide, density, ph, sulphates and alcohol
are physicochemical in nature, while the output variable 'quality' is based
on sensory data.

### Features of interest
Fixed and Volatile acidity, alcohol, density and sulphates look interesting 
and are good candidates for further exploration

### Did you create any new variables from existing variables in the dataset?
Total Acidity, which is the sum of fixed and volatile acidity was created. Since
volatile acidity is significantly low, the distribution of total acidity closely
resembles that of fixed acidity.


# Bivariate Plots & Analysis

```{r echo=FALSE, Bivariate_Plots-Plot16}
ggplot(aes(factor(quality), alcohol),data=redwine)+
  geom_jitter(alpha=0.1)+
  ggtitle("Alcohol & Quality")+
  geom_boxplot( alpha = .5,color = '#1f77b4')+
  stat_summary(fun.y = "mean", 
      geom = "point", 
      color = "red", 
      shape = 21, 
      size = 3)+
  xlab('Quality')+
  ylab('Alcohol (%)')+
  theme(panel.background = element_blank())
```

Alcohol is the first thing that comes to mind when I think of wine, so I wanted
to see the relationship between the quality of wine and its alcohol content.
Red wine of higher quality seem to have more alcohol content. The relationship 
is not a perfect linear one because most the wine are of medium quality (5,6) 
and the alcohol content in quality 6 is more spread out than in quality 5. 
However, there are more instances of high alcohol winerated at a higher quality. 
Correlation is not particularly high (0.5). However, we can clearly see from
the boxplots that the average alcohol content goes higher as we go from mid 
to top quality wines.

```{r echo=FALSE, Bivariate_Plots-Plot17}
ggplot(aes(factor(quality), volatile.acidity),data=redwine)+
 geom_jitter(alpha=0.1)+
 ggtitle("Acidity & Quality")+
 geom_boxplot( alpha = .5,color = '#1f77b4')+
  stat_summary(fun.y = "mean", 
      geom = "point", 
      color = "red", 
      shape = 21, 
      size = 3)+
 xlab('Quality')+
 ylab('Volatile Acidity (g/dm^3)')+
 theme(panel.background = element_blank())
```

From the data dictionary, we know that high levels of volatile acidity is not 
the mark of good wine, as high levels gives it an unpleasant, vinegar taste.
We suspected during when we looked at the distribution of volatile acidity in
wines that higher levels might correspond to low quality wines. This plot seems
to confirm our intuition. High quality wine tend to have less volatile acidity, 
which makes intuitive sense. Correlation is not particularly high (-0.4), 
but we see from the box plots that high quality wines tend to have low volatile 
acidity.

```{r echo=FALSE, Bivariate_Plots-Plot18}
ggplot(aes(x=density,y=total.acidity),data=redwine)+
  geom_point(color='#FF7F0E', alpha=0.2) +
  ggtitle("Acidity & Density")+
  xlab('Density (g/cm^3)')+
  ylab('Total Acidity (g/dm^3)')+
  theme(panel.background = element_blank())
```

Next, I wanted to explore if high levels of total acidity in the wine had any
relationship to the density of water in it. From the plot, we see a clear 
linear relationship between acidity and density. The larger the acidity, the more 
density in the wine. Correlation (0.7)


```{r echo=FALSE, Bivariate_Plots-Plot19}
ggplot(aes(x=alcohol,y=density),data=redwine)+
  geom_point(color='#1F77B4', alpha=0.2) +
  ggtitle("Alcohol & Density")+
  xlab('Alcohol (%)')+
  ylab('Density (g/cm^3)')+
  theme(panel.background = element_blank())
```

Lastly, how does density of the wine correspond to the alcohol content in it.
Higher water density corresponds to lower alcohol content and 
vice versa, as can be seen from the linear relationship between alcohol and 
density. Correlation is not particularly high (0.5)

Although pH was not one of the original features of interest, it is interesting
that there is a linear relationship between pH and fixed.acidity

```{r echo=FALSE, Bivariate_Plots-Plot20}
ggplot(aes(x=fixed.acidity,y=pH),data=redwine)+
      geom_jitter(color='#FF7F0E', alpha=0.3)+
      ggtitle("pH and Fixed Acidity")+
      xlab('Fixed Acidity (g/dm^3)')+
      ylab('pH')+
      theme(panel.background = element_blank())
```


# Multivariate Plots Section

In the Bivariate analysis, we saw interesting relationships between quality, 
acidity, density and alcohol. We will now explore further by adding more than 
two dimensions to these plots.

```{r echo=FALSE, multivariate_Plots-Plot21}
ggplot(aes(x=density,y=alcohol),data=redwine)+
  geom_point(aes(color=total.acidity), alpha=0.8) +
  scale_colour_gradient(name = "Total Acidity", 
                    low = "red", high = "blue")+
  ggtitle("Alcohol - Density - Total Acidity")+
  xlab('Density (g/cm^3)')+
  ylab('Alcohol (%)')+
  theme(panel.background = element_blank())
```

In the bivariate analysis, we saw the relationship between alcohol and density.
The higher the alcohol content, the lower the density of the wine. I wondered
how acidity fit into this relationship. Adding total acidity as a color scale 
to the alcohol - density plot, we see that total acidity is higher in wine with 
high density and spread across wine with various levels of alcohol

```{r echo=FALSE, multivariate_Plots-Plot22}
ggplot(aes(x=density,y=fixed.acidity),data=redwine)+
  geom_point(aes(color=log10(citric.acid)), alpha=0.8) +
  scale_colour_gradient(name = "Citric Acid (Log)",
                      low = "green", high = "darkgreen")+
  ggtitle("Fixed Acidity - Density - Citric Acid")+
  xlab('Density (g/cm^3)')+
  ylab('Fixed Acidity (g/dm^3)')+
  theme(panel.background = element_blank())
```

Citric acid is found only in small quantities in wine. Can it tell us more when 
we look at it with the relationshop we know that exists between density and
acidity? Content of citric acid is low in wine with low total acidity and 
density. This is seen by the light green dots towards the bottom left and the 
concentration dark green dots towards the top right. Here we have taken the log
of the citric acid variable.Correlation of citric acid to fixed.acidity is 0.7

```{r echo=FALSE, multivariate_Plots-Plot23}
ggplot(aes(x=alcohol,y=density, color=factor(quality)),data=redwine)+
  geom_point(alpha=0.3) +
  ggtitle("Alcohol - Density - Quality")+
  xlab('Alcohol (%)')+
  ylab('Density (g/cm^3)')+
  geom_smooth(method = "lm", se = FALSE,size=0.8)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+
  theme(panel.background = element_blank())
```
                    
So far, we have seen a few very interesting relationships between variables.
Our outcome variable is quality. Given the linear relationship between density
and alcohol, can encoding quality as color onto this plot tell us more about
these three variables?
Here, we see that most of the low rated wines are concentrated in the 
high density-low alcohol area (top left). the high rated 
wines are found more in the low density high alcohol are (bottom right).
When we draw regression lines for each quality category, we see that the lines
for the lower quality categories are towards the left and have a lower slope
as opposed to the higher quality lines, which are steeper and to the right.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = quality), data = redwine) +
  geom_bar(fill="#AEC7E8", stat='count') +
  labs(x='Wine Quality', y='Count') +
  ggtitle("Distribution of Wine Quality") +
  theme(panel.background = element_blank())
```

### Description One
This plot shows the distribution of wine by their quality, the output variable
of this dataset. As can be seen, the distribution is middle heavy, which means
that most wine are of quality 5 and 6 on the scale of (1-10). We do not have 
the information of wine across the quality scale. Hence we are limited in some 
of our inferences becausewe are trying to infer with only few of the low and 
high rated wines.


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x=density,y=fixed.acidity),data=redwine)+
  geom_point(aes(color=log10(citric.acid)), alpha=0.8) +
  scale_colour_gradient(name = "Citric Acid (Log)",
                      low = "green", high = "darkgreen")+
  ggtitle("Fixed Acidity - Density - Citric Acid")+
  xlab('Density (g/cm^3)')+
  ylab('Fixed Acidity (g/dm^3)')+
  theme(panel.background = element_blank())
```

### Description Two
The larger the density of the wine, the greater its acidity. We understand more
about the wine when we look at how this relates to citric acid, which is found 
only in small quantities in wine. Lower levels of fixed acidity and density is 
associated with lower levels of citric acid in the wine. Relationship of citric
acid is more stronger with fixed acidity (which intuitively makes sense). The
correlation between these two is (0.67)


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x=alcohol,y=density, color=factor(quality)),data=redwine)+
  geom_point(alpha=0.3) +
  ggtitle("Alcohol - Density - Quality")+
  xlab('Alcohol (%)')+
  ylab('Density (g/cm^3)')+
  geom_smooth(method = "lm", se = FALSE,size=0.8)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))+
  theme(panel.background = element_blank())
```

### Description Three
Higher water density corresponds to lower alcohol content and 
vice versa. High quality wine tend to be less dense and high in alcohol %.
Low and average rated wines tend to be high in density and low in alcohol.
This is seen by the concentration of such wines to the top left of the plot.
As we see from the regression lines of the wine quality categories, the lines
for the lower quality categories are towards the left and have a lower slope
as opposed to the higher quality lines, which are steeper and to the right.
This inference would have been stronger if we had more instances of wine
which are ranked low and high.

------

# Reflection

After exploring the Red Wine dataset, several interesting relationships were
discovered - The distribution of the features, relationship between features -
notabily those between acidity and density, acidity and alchohol and alcohol 
and acidity to quality. The relation between alcohol and density, acidity, 
density and traces of citric acid, residual sugar etc. gives us a good picture
of what constitues wine and how these features of wine relate to each other.

Ideally, such a dataset has the potential to tell us how these features relate
to the overall quality of the wine. Although we were able to make a few
inferences about quality, the fact that most wine were of medium quality didnt
give strenght to these inferences. This was a challenge in this analysis.
With more data, we would be able to explore these relationship in detail and
with more confidence.

With more data, I would also be confident to perform a regression analysis 
using the input features to predict the output variable - quality.


## References
http://www.calwineries.com/learn/wine-chemistry/acidity
https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt
